\def\title{CRAB: A Novel Clustering Score using Clustering with Rivals and Buddies for Unsupervised Learning}
\def\author{Allen Choi}
\def\degreemonth{June}
\def\degreeyear{2026}
\def\degree{Master of Science in Statistics}
\def\numberofmembers{3}
   \def\chair{Kelly Bodwin, Ph.D. \linebreak Thesis Advisor}
   \def\othermemberA{Alexander Dekhtyar, Ph.D. \linebreak Professor of Computer Science and Software Engineering}
   \def\othermemberB{Kevin Ross, Ph.D. \linebreak Professor of Statistics}
\def\keywords{Clustering. Unsupervised Clustering. Machine Learning.} % Make sure to end this in a . (i.e. Some, Thing, Interesting.)

\def\abstract{%
Unsupervised clustering algorithms today are used across a wide variety of fields
such biology, engineering, and industry in order to classify observations into groups where
labels are not provided. This can provide important latent information regarding the observations within
groups, as well as insight regarding the groups themselves. In order to judge the optimal number
of clusters for an unsupervised clustering algorithm, many methods exist such as the elbow method and Silhouette
Score; however, these methods come with drawbacks and are not necessarily flexible across many
unsupervised methods. We present a novel clustering score framework relying on a resampling-based approach
that classifies pairs of points as either "rivals" or "buddies", constructing a score for a given number
of clusters. We then present results demonstrating the properties of this metric on various simulated and
real datasets, measuring the performance of the algorithm as well.
}

\def\tocpagelegnth{1} % Number of pages your Table of Contents is (used to set correct page numbers for List of Tables and List of Figures)
\def\totflag{1} % 1 if you have at least 1 table, otherwise 0 (adds Table of Tables page)
\def\tofflag{1} % 1 if you have at least 1 figure, otherwise 0 (adds Table of Figures page)
